Assessment of forestry road network projection methods using the “roads” package in R

### 2022-07-18 analysis/paper/paper.Rmd
bring in all of David's text

# Introduction

Here is a citation [@Marwick2017]
Global biodiversity is being lost at an unprecedented rate as a result of human activities. One of the major world-wide threats to biodiversity is increasing infrastructural development (Benítez-López et al., 2010). Land transformation has resulted in the loss and fragmentation of habitats in many different ecosystem types worldwide (Benítez-López et al., 2010; Sanderson et al., 2002). Roads and other linear infrastructure, in particular, are key components of landscapes as they have become a major cause of habitat loss, fragmentation, and degradation and are ubiquitous in most landscapes around the world (van der Ree et al., 2011). Although linear infrastructure is important for society because it provides connectivity for people, it also exerts significant negative effects on adjacent habitats, wildlife populations, communities, and ecosystems (van der Ree et al., 2011).

# Methods
## Study Area
The Revelstoke timber supply area (TSA 27), in British Columbia, Canada, was selected as the main study site for this paper. With timber harvesting being the dominant resource extraction activity in this region, much of Revelstoke TSA’s total road network is composed of forestry resource roads which were built solely to gain access to forest harvest cutblocks; this makes for a suitable location to test our forest road projections which build networks using the same goal of connecting cutblocks. With approximately 55% of Revelstoke TSA’s area being non-forested land (alpine, lakes, swamp, brush, rock, etc) (Brown & Hachey, 2010), the TSA also provides an ideal location to observe the behaviour of the projections across a landscape which presents multiple barriers to road development. The projections were tested in various TSAs across the province with similar patterns being found in comparable mountainous landscapes where forestry is the dominant resource extraction activity (such as Arrow and Cranbrook TSAs). Revelstoke TSA contains wildlife species that are at risk due to their declining populations, such as members of the southern mountain woodland caribou herds (Brown & Hachey, 2010), and therefore is an example of a landscape where these projections could be used to manage critical wildlife habitats. The British Columbia TSA boundary shapefile layer was downloaded from the British Columbia Data Catalogue (Government of British Columbia, 2021a).

# Results

@@

# Conclusion

@@

>>>>

# Introduction[a]

Global biodiversity is being lost at an unprecedented rate as a result of human activities. One of the major world-wide threats to biodiversity is increasing infrastructural development (Benítez-López et al., 2010). Land transformation has resulted in the loss and fragmentation of habitats in many different ecosystem types worldwide (Benítez-López et al., 2010; Sanderson et al., 2002). Roads and other linear infrastructure, in particular, are key components of landscapes as they have become a major cause of habitat loss, fragmentation, and degradation and are ubiquitous in most landscapes around the world (van der Ree et al., 2011). Although linear infrastructure is important for society because it provides connectivity for people, it also exerts significant negative effects on adjacent habitats, wildlife populations, communities, and ecosystems (van der Ree et al., 2011).
Resource extraction, such as forest harvesting, is the main driver of road construction in much of Canada’s forested area. In British Columbia, over 600 000 kilometers of resource roads make up the vast majority of the province’s total road network (83%)  (Environmental Reporting BC, 2018; Forest Practices Board, 2015). Over 75% of resource roads in the province were built by the forest industry (Forest Practices Board, 2015). Resource extraction activities provide jobs and renewable raw materials (Canada, 2013b), but they also can create threats for wildlife. For example, the main threats to woodland caribou, a species at risk in Canada, are human disturbances such as forest harvesting, oil and gas exploration/extraction, and road networks. These particular disturbances fragment their habitat, and create open areas and extensive young forests that attract other species such as moose and deer which in turn, attracts an increased number of predators, such as wolves and even poachers (Canada, 2013a). Woodland caribou are often considered to be an umbrella species, which means that protecting and restoring their habitat would benefit other species also dependent on old forests, such as fishers, pine martens, and many species of birds (Mulvihill, 2018). 
With certain wildlife, such as woodland caribou, being highly sensitive to road development and use, there is a need to spatially simulate future roads in order to test the implications of alternative recovery policies and resource development scenarios. Landscape models are mathematical simulations designed to represent at least one landscape pattern-process relationship of interest. Landscape models can be great ways to make predictions about specific scenarios that allow ecologists and land managers to assess alternative management scenarios or to assess the potential impact of anthropogenic change scenarios (climate or land use change) (McGarigal, n.d.). Simulating the effects of disturbances, such as forestry road development, helps to broaden the scope of models, especially for species with complex ecological requirements who reside in disturbance-prone landscapes that undergo dynamic changes which may substantially alter habitat quality (Whitman et al., 2017). Land managers are often responsible for multiple species and populations distributed over large regions and often rely on models to determine habitat suitability (Whitman et al., 2017). 
Forest simulation models generally either fall into two categories: deterministic and stochastic (Boychuk et al., 2009; Helmes & Stockbridge, 2011). Stochastic simulation approaches use probability distributions in combination with random number generators to determine events (He & Mladenoff, 1999). Despite the importance of roads on the environment, previous spatial stochastic harvest simulation models rarely include forest road network development in their projections. The “roads” package in R aims to fill this gap in landscape modelling by projecting large-scale forest road networks based on future harvest cut blocks, an existing road network, and a road cost surface. This paper uses the “roads” package to run forestry road network simulations in British Columbia from approximately 1990 to present and compares the results to the observed modern-day forest road networks in order to assess the performance of the package and to showcase its usefulness within forest simulations.

(Introduction round 2: road context)
Several papers have previously introduced forest road network projections (Anderson & Nelson, 2004; Clark et al., 2000; Dean, 1997) however many of these methods are complex with many input parameters and are not readily available for public use. These previous projections have mainly been developed in order to facilitate operational forest road network planning and therefore require precise data and many input parameters in order to obtain the most effective network. These projection methods are therefore not very applicable for projecting forest roads over large-scale landscapes. On the other hand, many spatial stochastic forest management simulations haven’t included forest road development at all (citations). One unpublished QGIS plugin by Clement (2021) is available for projecting forest roads and is also included in the analysis of this paper (GitHub - Klemet/ForestRoadNetworkPluginForQGIS: A plugin for QGIS that is able to construct a forest road network based on a cost matrix, polygons to reach, a current road network, the dijkstra algorithm and 3 network-creation heuristics.). There is therefore a need in large scale landscape simulations for a tool that can adequately project forest road development with less complexity and minimal input parameters that can easily be incorporated into other complex simulations in large scale landscapes. As many ecological forecasts use increasingly complex methods and data, forecasting tools (e.g., models) should also be scalable and leverage existing models (Bodner et al., in press) (“Bridging the divide between ecological forecasts and environmental decision-making”).  
The goal of our model is to provide simple, accessible, and usable forest road network projections that can easily be applied in the context of large-scale landscapes and wildlife impact simulations, and can also be applied as an element of more complex models. Our model will be openly accessible and uses minimal input parameters (an existing road network, harvest cutblocks, and cost surface) which makes it easily transferable to many forest resource dominated landscapes. Here, we evaluate its accuracy at simulating forest road development across a large landscape by recreating the network starting from approximately 1990 and comparing it to the present day observed forest road network. In order to compare the projected and observed road networks, we used various metrics that are similar to those used in wildlife impact studies to compare the projected and observed landscapes both spatially and aspatially. 
In the first section of the paper (methods), we introduce the study area used to assess the accuracy of the projection, we describe the forest road projection methods, the data required, and the various metrics used in order to analyze the similarities and differences between the projected and observed forest road networks across the landscape. Next, we present and discuss the results of our projections across the landscape and compare the metrics spatially and aspatially.

# Methods
## Study Area
We chose the Revelstoke timber supply area (TSA 27), in British Columbia, Canada as our study area because forestry is the dominant resource extraction activity in this region and much of the TSA’s total road network is composed of forestry resource roads built to gain access to forest harvest cutblocks. This allows us to project road building based on forest harvest data. With approximately 55% of Revelstoke TSA’s area being non-forested land (alpine, lakes, swamp, brush, rock, etc) (Brown & Hachey, 2010), the TSA also provides an ideal location to observe the behaviour of the projections across a landscape which presents multiple barriers to road development. The Revelstoke TSA is in the interior-wet belt of the province and its forests are dominated by Englemann spruce (31%), Hemlock (23%), Cedar (22%), and Douglas-fir (18%). The TSA contains wildlife species that are at risk due to their declining populations and are believed to be negatively affected by road development, such as members of the southern mountain woodland caribou population (Brown & Hachey, 2010). The Revelstoke TSA is an example of a landscape where road projections could be used to manage critical wildlife habitats. The British Columbia TSA boundary shapefile layer was downloaded from the British Columbia Data Catalogue (Government of British Columbia, 2021a).

## Road network projection methods
Various optimization and simulation techniques have been used for projecting forestry roads into the future and across landscapes, however, we focused on the simulation methods that link multiple harvest units (a.k.a cutblocks) to an existing road network for strategic purposes. Dean (1997) described this road projection problem as the multiple target access problem (MTAP), where the objective is to find the minimum-cost locations for new roadways that link forested stands or landings (i.e., log deck) to the existing road network. We used three methods for solving the MTAP in the roads package. These methods represent a gradient of computational complexity and operational detail, starting with the least complicated or the snapping approach, followed by the least cost paths approach (LCP) and finally, the minimum spanning tree approach (MST). In the remainder of this paper, we will focus on the LCP and MST approaches given their assumptions are well aligned with how forestry professionals would strategically design road networks.
The LCP approach solves the MTAP by decomposing the problem into several smaller problems called single target access problems (STAP). For each STAP, a least cost path from a single source to sink was solved using Dijkstra’s shortest path algorithm (1959). Dijkstra’s algorithm minimizes the distance between a landing and the current road network, subject to costs or penalties incurred from road construction and travel (the details of this process are described later in the section “weighted graph”). Each STAP was then solved sequentially, with the newly projected roadway updating the existing road network and subsequently the costs of road construction. This method was first developed by Anderson and Nelson (2004) and provides a feasible solution to the MTAP, however the projected network is sensitive to the ordering of the STAP. They recommended a heuristic for connecting the landings and finding optimal branch locations to deal with this sensitivity.
The MST approach extends the LCP approach by providing a heuristic for linking landings or evaluating road branching opportunities. Dean (1997) showed that heuristics with branch evaluation were able to reproduce 80% of the observed road locations in their study area. Our MST approach has three steps towards evaluating branch locations while projecting roads: 1) estimate the least cost paths that link each of the landings and the least cost paths that link each landing to the existing road network, 2) creating a weighted graph where each node is either a landing or point on the existing road network and each edge or arc is weighted by the costs estimated in the previous step and 3) solving the minimum spanning tree for this graph. The minimum spanning tree was solved using Kruskal’s algorithm (1956) where every node in the graph was linked to form a tree such that the sum of edge weights is minimized, and the links do not form a loop. This method was most similar to Clark et al. (2000) approach which used a minimum spanning tree to project simple road networks during harvest scheduling. An important difference was that they based the edge weights on spatial Euclidean distance, whereas the MST approach used the costs estimated from least cost paths. 
        A projection created with the Forest Road Network Plugin for QGIS by Clement (2021) was also created within QGIS and added to the analysis in order to see how it compared with our projection methods. All the default settings such as skidding distance, etc. were left as they were within the QGIS plugin.

## Weighted graph
A mathematical weighted graph consisting of nodes and edges was used to parameterize either Kruskal’s or Dijkstra’s algorithms. When solving least cost paths using Dijkstra’s algorithm, each node in the graph corresponded to a pixel in a raster of our study area (100m x 100m resolution). Whereas, in the graph used for solving the minimum spanning tree via Kruskal’s algorithm, each node represented either a landing or a point on the existing road network. Therefore, a node can be represented by any one of three attributes in the forest road network: (i) a landing, (ii) a point on an existing road, or (iii) an intermediate point. The landing nodes represent log decks or sources of timber and the point on the existing road is the sink where the timber will end up. The intermediate points are plausible locations to construct a road and are thus free of barriers like major water bodies. The edges connecting the nodes follow a “queens” case or allow a maximum of eight possible edges for a given node. Thus, our methods utilize a grid pattern which does not support curvature or switchbacks in road design as required in operational road design, thus we ascertain that our road projection methods are for strategic modelling purposes only. Edges are assigned weights which penalize the algorithm to account for various impedance and road construction costs. When solving least cost paths, a cost surface raster was used. The pixels of the cost surface raster corresponded to nodes in the weighted graph and thus the edge weight was the average cost between the two nodes or pixels. When solving the minimum spanning tree, the edge weight corresponded to the total cost of the least cost path between the two nodes.

## Data
To spatially define the cost of constructing and traveling on a forestry road we developed a cost surface raster. We used several spatial raster layers describing the roading cost structure, barriers to road development and road impedance. For intermediate nodes, the roading cost structure followed the Interior Appraisal Manual of BC (source) where the cost per kilometer of road was a linear function of slope. Separate costs for river and pipeline crossings were then added to the roading cost. Barriers for road development masked the cost surface raster and these included the major water bodies of BC. Lastly, to coerce the algorithm to follow existing roads, all nodes representing the existing road network were assigned the smallest cost (Table 1). The cost surface for Revelstoke TSA consisted of 1703936 pixels with a resolution of 100 meters by 100 meters.

Table 1. Definitions of road building costs for building the cost surface. 
Cost Type
	Road Attribute
	Value ($ per km)
	Source
	Barriers of road development
	Major lakes
	infinity
	Major water bodies of BC
	Impedance
	Existing roads
	0.01
	Assumed
	Road Construction Cost Structure

	Pipe crossing
	1911
	Interior Appraisal Manual
	River crossing
	2130
	Interior Appraisal Manual
	Base cost
	16178 + 504*slope
	Interior Appraisal Manual

The road network used was a combination of the digital road atlas (DRA) (Data BC, 2021) and forest tenure road (FLNRORD, 2021b) datasets from the Government of British Columbia’s Data Catalogue. The existing road network inputted into the model was created by filtering out the forest tenure roads with an award date greater than 1990. Because the objective of the roads package is to project forest resource roads, the DRA roads were left untouched for the existing 1990 road network. The projection results would be compared with the observed modern road network which was simply a complete and unfiltered combination of the DRA and Forest Tenure road datasets, representing the most complete and up-to-date road network for the area.   
Cutblock information was obtained from the Harvest Areas of BC (Consolidated Cutblocks) Government of British Columbia’s Data Catalogue (Government of British Columbia, 2021b). Cutblock information is compiled from provincial Forest Cover, the RESULTS Reporting system, Forest Tenure applications, and satellite imagery using change detection processes. The dataset does not exclude non-harvested ‘reserved’ areas, and satellite image change detection may not always represent the entire cutblock area (Government of British Columbia, 2021b). Forest harvest cut blocks with harvest years greater than 1990 were filtered and used as targets for the projections.  
Analysis
Projections were run in R (R Core Team, 2020) within the Revelstoke TSA. The forest road network projections were run using two different resolutions of the cost surface in order to assess how sensitive the results were to pixel size; the finer resolution used cost surface pixels of 100 meters by 100 meters and was the focus of this study, whereas the coarse resolution used an aggregated cost surface with pixels of 1000 meters by 1000 meters. The least cost path method yields projections that are obviously less realistic than the minimum spanning tree method (Figure 2), so we focussed on assessing the performance of the minimum spanning tree method and investigated sensitivity to sampling types (regular and random) and sampling densities (0.000001 for low and 0.0001 for high) (Figure 2).     
        In order to assess the performance of the projections, the following metrics were used to compare the projected forest road network results with the observed network. Road presence is a simple measure of whether or not roads are present in each pixel using the cost surface as a template. Road disturbance footprint was created as done by Ibisch et al. (2016) where a 1 km buffer was applied to all roads. Forestry disturbance footprint was created by buffering 500 metres around all roads and cutblocks. This metric relates to the anthropogenic disturbance metric used for boreal caribou (Johnson et al., 2020), and can be calculated using the caribouMetrics R package (Endicott et al.,in prep). The 100m x 100m resolution cost surface was used as the template to derive the forestry disturbance footprint metric values. Road density is an important metric that is often used when studying anthropogenic impacts on wildlife (Beazley et al., 2004). Road density is represented in meters per hectare (m/ha) and was calculated using the rasterizeLineDensity function within the caribouMetrics package (Endicott et al., in prep). This was done by intersecting the roads with the cost surface raster grid, calculating the length of road in each grid cell and dividing by the grid cell resolution. Distance to nearest road assigns each cell within the TSA a distance (meters) to the nearest road segment. The metric can be calculated using the “dist_to_nearest” function in the “starsExtra” package (Dorman, 2021), but this implementation can be prohibitively slow. We have implemented a faster moving window method in the XX function of the roads package.
        Using the metrics described above, we assessed the aspatial accuracy of the projections by calculating the mean values of the projection metrics rasters and then compared them with the means of the observed metric rasters. This was done at an overall TSA level (where means were taken from across the entire study area) and at a cutblock level (where metric rasters were masked by the cutblock polygons and the means were taken only from within these areas).   
        The projections were assessed for their spatial accuracy in two different ways. First, difference maps were created by subtracting the road presence rasters of the projected networks with the observed network in order to further highlight discrepancies between the networks. The observed and projected road network shapefiles were then overlain on the difference maps and regions of major discrepancies were investigated using Google satellite imagery in GIS (QGIS) to gain a better understanding of what was happening on the ground. Secondly, a roaded/roadless matching analysis using the three binary metrics (road presence, road disturbance footprint, and forestry disturbance footprint) was performed. This matching analysis compared the three metric rasters for all projection methods with the observed metric rasters and calculated percentages of the TSA which matched and didn’t match. All pixels in the Revelstoke TSA for the projected networks         fell into 5 categories when compared with the observed raster: Roaded prior to 1990 is the roaded area of the existing 1990 road network that was inputted into the models to project future forestry roads; agree roadless is the roadless area where pixels between projected and observed match; agree roaded after 1990 is the roaded areas where projected forestry roads match with observed forestry roads; project roaded & observe roadless is areas where the projection was roaded but where the observed network was roadless; and finally project roadless & observe roaded represents areas where the projection was roadless but the observed network was roaded. Using these 5 categories, the landscape within the Revelstoke TSA was categorized for each projection method and summarized (figure 3).

# Results
## Aspatial accuracy
Various projection methods provided more accurate aspatial results than others when compared with the observed metrics depending on the metric used. The forestry disturbance footprint metric produced means across all projections that were closest with observed values (Figure 3). Projections of road disturbance footprint and distance to road were also reasonably accurate across the whole landscape, with distance to road being less accurate within cutblocks. Road presence and road density were less accurate and much more sensitive to assumptions about the distribution of roads within cutblocks (i.e. sample density). The road presence metric is sensitive to small differences in road location while the road density metric is sensitive to variations in the curves of tightly packed roads within grid cells that cannot be modeled using this algorithm. Road presence and road density were also the most sensitive to pixel size (reference appendix table for coarse resolution projections). The projection using the Clement’s (2021) QGIS plug-in also yielded quite accurate results, especially at an overall TSA level.

## Spatial accuracy
        Difference maps helped to highlight areas of discrepancy between the projections and the observed forestry road networks. In areas where the quality of the cutblock and road data was high, the projections remained within a reasonable vicinity of the observed roads and also realistically remained within valley bottoms without cutting across natural barriers such as mountains (figure 4a). Areas of greatest discrepancy often suggest problems with available forestry road or cutblock data. There are some areas in Revelstoke, for example, where there are mapped cutblocks but no mapped access roads, even though satellite images clearly show roads (Figure 4b). There are also instances where areas were likely misidentified as cutblocks (Figure 4c). This is likely due to difficulty distinguishing cutblocks from other unforested areas such as scree slopes and meadows using remote sensing methods.
Figure 5 shows the percentage of matching pixels by category between observed and projected rasters. The percentage of roaded prior to 1990 was the same across all projection methods for the respective metrics because the same 1990 road network was used for all projection as the input for the existing road network. The agree roadless category represented most of the landscape for all projection methods. The subsequent 3 categories (Agree roaded after 1990, project roaded & observe roadless, project roadless & observe roaded) represent the areas where change occurred following 1990.
With the road disturbance footprint metric, the agree roaded after 1990 represented the largest percentage of the landscape which underwent change after 1990. The project roaded & observe roadless category represented higher percentages of the landscape than the project roadless & observe roaded, especially when using a higher sampling density, which shows that there were more areas where the model over projected roads than under projected roads. The small percentage of project roadless & observe roaded across all projection methods shows that there were very few observed roaded areas that the projections did not address.  
Using the forestry disturbance footprint metric, the post-1990 roaded pixels lined up very well. Nearly 70% of all post-1990 projected roaded pixels matched with the observed network when using this metric.  Project roadless & observe roaded very slightly made up a higher percentage of the landscape when compared to project roaded & observe roadless (by less than half a percent across all projection methods). 
When using the fine resolution road presence raster to compare with the observed road presence raster, very few of the post-1990 pixels align. Nearly all post-1990 roaded pixels did not match between observed and projected. The sampling density used determined whether there were more project roaded & observe roadless pixels (higher sampling density) or project roadless & observe roaded pixels (centroid and lower sampling density) amongst the post-1990 roaded pixels.

@@

Through assessing the performance of our model by comparing its spatial and aspatial
accuracy with the observed forest road network following 1990, we are satisfied that the projections provide adequate results given their intended use, and that it is a useful tool for projecting forest road networks across a large landscape with minimal input parameters. However, for applications where the exact location of roads is important (like operational road engineering purposes), or for projecting impacts on wildlife that respond to roads at a very fine spatial scale then this is not an appropriate model. The model does, however, have potential to be a useful tool for predicting impacts on wildlife across large landscape simulations and could easily be incorporated as an element into more complex models that take into account other landscape phenomena such as wildfires, climate change, etc. Because this model does not take timescales into account, it could be a useful post-processing tool as an element of more complex models as it can be applied retroactively to landscape simulations.
        Many of the major discrepancies between the projections and the observed network had to do with potential issues with the available data; these discrepancies should be treated with caution as they are not necessarily the fault of the projections themselves and rather with issues in available data. These potential issues were identified through the use of the difference maps paired with recent satellite imagery that allowed us to better understand what was happening on the ground. Often major discrepancies were related to cutblocks that were likely falsely labelled (most often through LANDSAT change detection) in places where there weren’t actually cutblocks. These discrepancies had a negative influence on the overall spatial and aspatial accuracy of the projections. In order to get more accurate road projections, one could simply remove the cutblocks derived from LANDSAT as these were often the cutblocks which caused issues, however, these cutblocks were left in as a major goal of this assessment was to create road projections while adjusting input parameters as little as possible. Considering how this model will be used to simulate forest roads moving into the future, it is unlikely that potential future cutblocks will be mislabelled as they were in our situation and this would not be an issue. Another potential issue was with roads not being included in the data set that were seen in the satellite imagery. This issue relates to results from the study by Poley et al (in review) that demonstrated how road network datasets are almost universally incomplete and that many low-use roads (such as forest resource roads) can be seen within recent satellite imagery but are not included in any readily available road datasets. Mapping of roads is often especially lacking in more rural, forested, and mountainous regions where roads are often unpaved (Prendes et al., 2017) with the Auditor General of British Columbia (2017) suggesting that 100 000 km of resource roads are missing from publicly-available road network maps (inspired by Adam Ford’s doc). This means there were areas where the projections built a road, causing a discrepancy, but in reality there was actually an observed road there as seen in the satellite imagery. These discrepancies also negatively impacted the spatial and aspatial accuracy results when they shouldn’t have and should be treated with caution. It is possible that some roads seen in the satellite images have since been closed which could be a reason for them being absent in the roads data sets. Again, this would not present an issue when predicting future impacts on landscapes, only when assessing the model accuracy in this context. Having said this, the projections were deemed most accurate in areas where the quality of the cutblock and road data was high and consistent with what was seen in recent satellite images.
Although this model works well in an area like Revelstoke where the road network is located in a region dominated by forestry roads, it does not work well in landscapes where roads are developed based on other targets outside of cutblocks (where the intent to build the road is not to achieve the least cost path; aka seismic lines). Fort Nelson TSA (TSA 8), for example, contains a large amount of non-forestry roads which are related to seismic lines, oil and gas extraction, etc. Because of this, forestry roads make up a much smaller percentage of the overall road network. In landscapes like this, it does not make sense to use the forestry road projections as many roads are influenced by many different factors. The model was attempted in the Fort Nelson TSA however the resulting road network (Figure A-1) did not accurately portray the observed network like it did in the Revelstoke TSA. However, other resource extraction activities could likely be incorporated into this model relatively easily. The same concept could be used where mining or oil locations could be used as target locations instead of cutblock areas. 
 It is also worth noting that although the model can project road networks fairly accurately across a large landscape, it may differ from the way in which roads are planned in real life. The model will simply build the roads in the least costly locations based on the cost surface, however often in the real world, roads must be planned around ownership and administrative boundaries. In BC, forestry roads are developed by forest licensees under a timber harvesting permit (Government of British Columbia, 2021c). 
Certain projection methods were more accurate at projected some metrics compared to others. Similarly, certain metrics were sensitive to pixel size (road density, road presence) and others were not. Depending on what the goal of running the projections is, then the projection method and resolution may or may not play an important role. If you would like to predict road density as accurately as possible, then running the projections using as fine a resolution of cost surface as possible is the best option. If you’re wanting to only predict roughly the location of the roads/save computing time and are more interested in a larger area, then a coarse resolution may suit these requirements perfectly fine. The tradeoffs of using a coarser resolution of cost surface may be less extreme when projecting roads across a flatter landscape compared to a mountainous landscape like was used for this paper since there would likely be less variability in cost across larger areas.  
     With caribou being one of the key considerations when managing some forestry areas (like in British Columbia TSAs, including Revelstoke TSA) (Government of British Columbia, 2021), the accuracy at which “roads” can project this forestry disturbance footprint metric could allow for this package to be a truly beneficial tool for assessing the impacts towards at-risk caribou populations when looking at various forest harvest scenarios. Because caribou are also considered an umbrella species within forests (Drever et al., 2019), projecting forestry impacts on their habitats may also provide valuable insight towards other key species as well. With the input of future cutblocks and the projections of forest road networks to these cutblocks, the results from the projection can provide a great outlook at the impacts of forestry scenarios on woodland caribou habitats using the forestry disturbance footprint metric. The distance to nearest road metric appears to be a very interesting metric to consider as well when looking specifically at larger areas. Using this metric to make projections at larger scales, such as at a TSA level, may provide very beneficial information about forestry impacts especially in a wildlife context when assessing the influence of forestry on various species with different tolerances to disturbance.  
Further improvements could be made that would increase the accuracy of the model. The cost surface used for this paper was developed at a provincial scale, however, a higher resolution cost surface with more detail could further improve the accuracy of the projection results. A finer resolution cost surface with more detail will ultimately create a more realistic road network and would possibly allow for more accuracy within cutblocks (where fine micro-topography, water tables, etc. will play a role in road development). The downside being that it would also increase computing time and may not always be necessary depending on the application of the model, especially when using the model for a large-scale landscape. Adding too much complexity to the cost surface may also take away from the goal of this model to be simple and easy to use, however, the algorithm itself would work on a very fine and detailed cost surface. Along with increased computing time, a key limitation of using this model on a finer cost surface would be that eventually cost pixels would need to have directionality associated with them, which is something this model does not take into account. The accuracy of comparing the projections to the observed network could also simply be improved by using higher quality/more up to date road and cutblock data. 
Assessing the performance of this model also presented some unique challenges with comparing two networks with each other across a landscape. We needed to carefully think about what we were trying to achieve. The goal wasn’t to present a model that matched the observed road network pixel by pixel, but instead was to simulate disturbance across a large-scale landscape to an acceptable level while still being a relatively simple and straightforward tool. It’s for this reason why metrics were chosen that are commonly used in road impact assessments on wildlife; such as the road density and forestry disturbance footprint metrics. Once the vector networks were converted into these raster layers, it also facilitated comparing observed versus projected as we could compare the means across the TSA and study where pixels matched and didn’t match. Incorporating these metrics into landscape level simulations is also a realistic goal for this tool (ie projecting forest road density across a landscape) so a goal was to therefore see how well the model did at simulating these metrics when compared to the observed landscape.   
Many spatial stochastic models don’t include road projections at all in their simulations (Bergeron et al., 2017; DeLong & Tanner, 1996; Naderializadeh & Crowe, 2020; Perera & Cui, 2010). Our model could be relatively easily incorporated, or applied retroactively, to similar spatial stochastic simulations such as these. For large-scale projects such as the Western Boreal Initiative (Environment Canada, date?) that simulate many complex factors across a landscape, our forest road projection methods could be incorporated as one element within the overall landscape simulations.

# Conclusion
The forest road network projection model presented here provides an effective way to project forest road networks that are suitable for large scale landscape simulations with relatively little effort. This model, however, is not suitable for strategic planning. The results from our projections in our study area overall matched up adequately with the observed present day forest road networks, with major discrepancies being attributed to issues in the available road and cutblock data. These discrepancies would not be issues in future landscape simulations, only when assessing accuracy. In areas where the quality of the data was high, projections matched well with the observed network. Our metrics were able to be projected effectively across a large landscape and could be useful indicators of impacts on wildlife habitats in future simulations. 
        This model was developed with the goal of being simple to use, readily accessible, and relatively accurate across a large landscape. With minimal input parameters, complexity, and computational power, one can project future forest road networks using this tool across large areas which can aid in the assessment of potential impacts of forestry disturbance on wildlife and habitat over a large spatial scale.  With land use activities transforming a large proportion of the planet’s land surface and becoming an issue of global importance (Foley et al., 2005), there is an increasing need to improve and build upon landscape level models in order to properly manage and protect remaining natural habitats. Land-use activities, primarily for agricultural expansion and timber extraction, have caused a net loss of ~7 to 11 million km2 of forest in the past 300 years (Foley et al., 2005). With only 7% of the planet’s roadless areas being larger than 100 square kilometers, there is a need to protect key areas which sustain key refugia for biodiversity from fragmentation and degradation (Ibisch et al., 2016). The introduction of this widely applicable and user-friendly forest resource road model is a step in the right direction for providing more tools to land managers and planners who are interested in assessing the potential impacts of road development over large landscapes.

@@

[a]A comment
